Adaption of the Chumbley Score to matching of bullet striation marks

### 2018-01-03 p2/p2.Rmd
.

Identification criteria like consecutively matching striae (CMS) as first seen by Biasotti 1959 \citet{biasotti} and later mentioned in \citet{chu2013} more often than not include subject bias and are error prone and as seen in the work of \citet{miller} the number of CMS may turn to out to be high even when the bullet is not fires by the same firearm. Parameters like cross-correlation factor as seen in the extensive comparisons made by \citep{aoas} seem to perform much better matching purposes.

@@

## The Chumbley Score Test

\hh{You need to distinguish between $x(t)$ and $x(t_1)$ - $x(t)$ is a spatial process for some $t$. $x(t_1)$ is the realization of the process in $t_1$. - Alternatively you could also follow the notation in Hadler et al on 'digitized toolmarks at a pixel level' - where our 'pixel' is the resolution of the confocal light microscope, i.e. a pixel corresponds to 0.645 microns.}

The Chumbley score algorithm takes input as two vectorized processes $x(t_1)$ and $y(t_2)$ which denote two sets of marks or striae. These marks or striae are potentially from two different bullets or two different toolmarks whose source needs to be identified as being same or different. The marks or striae are indexed by the pixel location where $t_1$ is for the first striae referred to as $x$ and $t_2$ is for the second striae which is referred as $y$. $t_1$ and $t_2$ need not be the same but are usually of similar lenghtsThe similarity is then judged by the algorithm on the basis of cross-correlation of a fixed and constant number consecutive pixels say $k$ taken from the two indexed marks $x(t_1)$ and $y(t_2)$ such that in theory $k$ remains smaller that length of the two striae or marks which is the same as $t_1$ and $t_2$. Depending on the what stage of the algorithm we are in, matching of different pixel lengths and locations is done, which at the end effectively compares all possible windows that would guarantee in quantifying the two marks or striae as coming from the same source or not, which also lets us assess the error rates by checking for a large number of cases.

@@

## Testing setup:
\hh{Using cross-validation setup to identify appropriate parameter settings for (a) signatures and (b) profiles directly}

@@

The marking made on bullets are smaller than toolmarks and is also less wider. The idea is to find out possible areas of error while adapting the score based method proposed for toolmarks.

Bullet signatures being compared at this time are from the Hamby 44 and Hamby 252 data, which have the best set of known-matched and known non-matches.

Bullet signatures are chosen by
1. Filtering out Land_id for Profiles from the Hamby 44 and Hamby 252 data and removing 
all NA values
2. run_id = 3 is chosen, Seems like the signatures generated from this run_id give the closest
match. Different run_id's have some different settings for generating the signatures
(The level of smoothing does not seem to be one of them)

The bullet signatures when generated already included the LOWESS smoothing.
Therefore, the coarseness factor is set to 1 while running the chumbley_non_random()
which generates the same_shift, different_shift, U-Stat and P_value parameters.

```{r }
library(kableExtra)
a27<- all %>% filter(wo==200, wv==30) %>% xtabs(data= ., ~signif+match)
knitr::kable(a27, digits=0, caption = "Window of Optimization 320 and Window of Validation 50", format = "latex" ,booktabs=TRUE)%>% kableExtra::kable_styling(latex_options = c("hold"))

```{r}
a27<- all %>% filter(wo==120, wv==30) %>% xtabs(data= ., ~signif+match)
knitr::kable(a27, digits=0, caption = "Window of Optimization 120 and Window of Validation 50", format = "latex" ,booktabs=TRUE)%>% kableExtra::kable_styling(latex_options = c("hold"))
```

```{r}
a27<- all %>% filter(wo==80, wv==30) %>% xtabs(data= ., ~signif+match)
knitr::kable(a27, digits=0, caption = "Window of Optimization 80 and Window of Validation 50", format = "latex" ,booktabs=TRUE)%>% kableExtra::kable_styling(latex_options = c("hold"))
```

## Signatures

@@

We used the adjusted Chumbley method as proposed in \citet{hadler} and implemented in the R package `toolmaRk` \citep{toolmark} on all pairwise land-to-land comparisons of the Hamby scans (a total of 85,491 comparisons) in the following manner:

```{r, echo=FALSE, results='asis'}
param_setting<- all %>% select(wo, wv) %>% group_by(wo, wv) %>% tally() %>% select(wo,wv)
#param_setting<- data.frame(matrix(param_setting, nrow = 16))
#t(param_setting)
#param_setting %>% tally() 
library(kableExtra)
knitr::kable(t(param_setting)[,1:17], digits=0, caption = "Overview of parameter settings used for optimization and validation windows for bullet land signatures.", format = "latex" ,booktabs=TRUE) %>% kableExtra::kable_styling(latex_options = c("scale_down", "hold"))

@@

@@

Figure \ref{fig:type1} compares nominal (fixed) type I error and actually observed type I errors for the parameter settings in table XXX. With an increasing size of the window used in the optimization step the observed type I error rate decreases (slighty). This might be related to the increasing number of tests that fail for larger window sizes, in particular for non-matching striae (see fig \ref{fig:missings}).

@@

```{r wo120, fig.width = 8, fig.height=4.5}
wo120 <- errors %>% filter(wo==120)
wo120 %>% gather(type, error, actual:beta) %>%
  ggplot(aes(x = wv, y = error, colour = factor(alpha)))  + 
  geom_line() + geom_point() + facet_wrap(~type, scales="free") +
  theme_bw()+ 
  theme(legend.position="bottom")
```

@@

Coming back to the database of Hamby-44 and Hamby-252 datasets, the run_id = 3 was used when applying the chumbley algorithm on the signatures. The run_id not only defines the level of smoothing but also signifies the chosen height at which the profiles were selected initially.
Another important aspect is the range of horizontal values (which is referred to as the y values in \citet{aoas}) in the signatures. Thes have already been pre-processed in the database to not include any grooves.

\hh{XXX ideas should go to the previous section on setting up the study; in this section we focus on results only XXX}
The idea therefore is to first use these raw values of the profile directly in the chumbley algorithm, and see how the algorithm performs for different coarseness values (smoothing parameter as referred in the function LOWESS used in the chumbley algorithm)

@@

total1<- errors_sig_c %>% ggplot(aes(x = coarse, y=total, colour=factor(alpha), shape="signatures")) + geom_point() + geom_line()  + geom_point(data= errors2 %>% filter(wv==30), aes(shape="profiles")) + geom_line(data=errors2 %>% filter(wv==30), aes(shape="profiles")) + theme_bw() +
  scale_colour_brewer(expression("type I error"~alpha), palette="Set2")

total2<- wo120 %>% filter(wv>10) %>% ggplot(aes(x = wv, y=total, colour=factor(alpha), shape="signatures")) + geom_point() + geom_line() +geom_point(data=errors2 %>% filter(wv>10) %>% filter(coarse==0.25), aes(shape="profiles")) + geom_line(data=errors2 %>% filter(wv>10)%>% filter(coarse==0.25), aes(shape="profiles")) + theme_bw()

@@

## Results of Chumbley score
- Nominal alpha value shows dependency on the size of the window of optimization
- Test Fail depends on whether known-match or known non-matches has predictive value
- Type II error least bad for window of validation 30 and window of optimization 120

>>>>

Identification criteria like consecutively matching striae (CMS) as first seen by Biasotti 1959 \citet{biasotti} and later mentioned in \citet{chu2013} more often than not include subject bias and are error prone and as seen in the work of \citet{miller} the number of CMS may turn out to be high even when the bullet is not fired by the same firearm. Parameters like cross-correlation factor as seen in the extensive comparisons made by \citep{aoas} seem to perform much better for matching purposes.

@@

# The Chumbley Score Test

The Chumbley score algorithm takes input as two vectorized processes $x(t_1)$, $t_1 = 1,2,...T_1$ and $y(t_2)$, $t_2 = 1,2...T_2$ which denote two sets of marks or striae. Here the processes are of the form $z(s)$ which is a spatial process for some $t$, while $z(s_1)$ is the realization of the process in $s_1$. The $t_1$ and $t_2$ can be better understood as equally spaced pixel locations for the two marks under consideration, where a 'pixel' is the resolution of the confocal light microscope. In the case for bullet signatures a pixel corresponds to 0.645 microns. These marks or striae are potentially from two different bullets or two different toolmarks whose source needs to be identified as being same or different. The marks or striae are indexed by the pixel location where $t_1$ is for the first striae referred to as $x$ and $t_2$ is for the second striae which is referred as $y$. $T_1$ and $T_2$ being the respective lengths of the markings, need not be the same but are usually of similar lenghtsThe similarity is then judged by the algorithm on the basis of cross-correlation of a fixed and constant number consecutive pixels say $k$ taken from the two indexed marks $x(t_1)$ and $y(t_2)$ such that in theory $k$ remains smaller that length of the two striae or marks which is the same as $t_1$ and $t_2$. Depending on the what stage of the algorithm we are in, matching of different pixel lengths and locations is done, which at the end effectively compares all possible windows that would guarantee in quantifying the two marks or striae as coming from the same source or not, which also lets us assess the error rates by checking for a large number of cases.

@@

## Testing setup

@@

The marking made on bullets are smaller than toolmarks and is also less wider. The idea is to find out possible areas of error while adapting the score based method proposed for toolmarks, using cross-validation setup to identify appropriate parameter settings for (a) signatures and (b) profiles directly

@@

Bullet signatures being compared at this time are therefore from the Hamby 44 and Hamby 252 data. The database setup and pre-processing system used for choosing the Bullet signatures are as described by \citet{aoas}. In order to choose the bullet signatures we first filter out Land_id for Profiles from the Hamby 44 and Hamby 252 data and remove all NA values. Then run_id = 3 is chosen as the signatures generated from this run_id give the closest match. Different run_id's have some different settings for generating the signatures.(The level of smoothing does not seem to be one of them)

The bullet signatures when generated by this process already includes a loess smoothing. Therefore, the coarseness factor is set to 1 while running the chumbley non random algorithm for comparing different optimization windows.The algorithm generates the same_shift, different_shift, U-Stat and P_value parameters which are then used to calculate the errors associated with different sets of window sizes.

We used the adjusted Chumbley method as proposed in \citet{hadler} and implemented in the R package `toolmaRk` \citep{toolmark} on all pairwise land-to-land comparisons of the Hamby scans (a total of 85,491 comparisons) with the pairwise sets for the comparisons given in the following table.

```{r param, echo=FALSE, results='asis'}
param_setting<- all %>% select(wo, wv) %>% group_by(wo, wv) %>% tally() %>% select(wo,wv)
#param_setting<- data.frame(matrix(param_setting, nrow = 16))
#t(param_setting)
#param_setting %>% tally() 
library(kableExtra)
knitr::kable(t(param_setting)[,1:17], digits=0, caption = "Overview of parameter settings used for optimization and validation windows for bullet land signatures.", format = "latex" ,booktabs=TRUE) %>% kableExtra::kable_styling(latex_options = c("scale_down", "hold"))

@@

Similar to signatures the run_id = 3 was used when applying the chumbley algorithm using the database setup given by \citet{aoas} of Hamby-44 and Hamby-252 datasets, on the profiles. The run_id not only defines the level of smoothing but also signifies the chosen height at which the profiles were selected initially. Another important aspect is the range of horizontal values (which is referred to as the y values in \citet{aoas}) in the signatures. These have already been pre-processed in the database to not include any grooves.

The idea therefore is to first use these raw values of the profile directly in the chumbley algorithm, and see how the algorithm performs for different coarseness values (smoothing parameter as referred in the function lowess used in the chumbley algorithm)

```{r }

#knitr::kable(a27, digits=0, caption = "Window of Optimization 320 and Window of Validation 50", format = "latex" ,booktabs=TRUE)%>% kableExtra::kable_styling(latex_options = c("hold"))

```{r}

#knitr::kable(a27, digits=0, caption = "Window of Optimization 120 and Window of Validation 50", format = "latex" ,booktabs=TRUE)%>% kableExtra::kable_styling(latex_options = c("hold"))
```

```{r confusion}
library(kableExtra)
a27<- all %>% filter(wo==200, wv==30) %>% xtabs(data= ., ~signif+match)
a27<- as.data.frame(a27)
a27$Type<- c("True Negative","False Positive (Type I)","False Negative (Type II)","True Positive")

a28<- all %>% filter(wo==120, wv==30) %>% xtabs(data= ., ~signif+match)
a28<- as.data.frame(a28)
a28$Type<- c("True Negative","False Positive (Type I)","False Negative (Type II)","True Positive")

a29<- all %>% filter(wo==80, wv==30) %>% xtabs(data= ., ~signif+match)
a29<- as.data.frame(a29)
a29$Type<- c("True Negative","False Positive (Type I)","False Negative (Type II)","True Positive")

knitr::kable(list(a27,a28,a29), digits=0, caption = "Confusion Table for different optimization window sizes with validation window size as 30", format = "latex" ,booktabs=TRUE)%>% kableExtra::kable_styling(latex_options = c("hold"), "striped", full_width = F) %>% group_rows("Size of Optimization Window = 200",1,5) %>% group_rows("Size of Optimization Window = 120", 6, 10) %>%  group_rows("Size of Optimization Window = 80", 11,14)
```

@@

Figure \ref{fig:type1} compares nominal (fixed) type I error and actually observed type I errors for the parameter settings in table \ref{tab:param}. With an increasing size of the window used in the optimization step the observed type I error rate decreases (slighty). This might be related to the increasing number of tests that fail for larger window sizes, in particular for non-matching striae (see fig \ref{fig:missings}).

@@

```{r wo120, fig.width = 8, fig.height=4.5, fig.cap="For optimization window size 120 and different validation window size how Actual and Type II (beta) errors vary for various chosen nominal alpha levels"}
wo120 <- errors %>% filter(wo==120)
wo120 %>% gather(type, error, actual:beta) %>%
  ggplot(aes(x = wv, y = error, colour = factor(alpha)))  + 
  geom_line() + geom_point() + facet_wrap(~type, scales="free") +
  theme_bw()+ 
  theme(legend.position="bottom")
```

@@

@@

total1<- errors_sig_c %>% ggplot(aes(x = coarse, y=total, colour=factor(alpha), shape="signatures")) + geom_point() + geom_line()  + geom_point(data= errors2 %>% filter(wv==30), aes(shape="profiles")) + geom_line(data=errors2 %>% filter(wv==30), aes(shape="profiles")) + theme_bw() +
  scale_colour_brewer(expression("Nominal\ntype I error"~alpha), palette="Set2")

total2<- wo120 %>% filter(wv>10) %>% ggplot(aes(x = wv, y=total, colour=factor(alpha), shape="signatures")) + geom_point() + geom_line() +geom_point(data=errors2 %>% filter(wv>10) %>% filter(coarse==0.25), aes(shape="profiles")) + geom_line(data=errors2 %>% filter(wv>10)%>% filter(coarse==0.25), aes(shape="profiles")) + theme_bw() +
  scale_colour_brewer(expression("Nominal\ntype I error"~alpha), palette="Set2")

@@

## Results of Chumbley score
- Nominal alpha value shows dependency on the size of the window of optimization
- Test Fail depends on whether known-match or known non-matches has predictive value
- Type II error least bad for window of validation 30 and window of optimization 120
- Type II error for profiles increases with the amount of smoothing and least for LOWESS smoothing coarseness value about 0.25 or 0.3.
- Profiles in comparison with Signatures seem to perform worse with total error either equal or more for all sizes of validation window.
-
